---
title: "BIOL7130 Experiment Report"
author: "Yasmin Roustaei"
date: "2 June 2021"
output:
  word_document: default
  html_document: default
---

##Title
Effects of mountain elevation on the genetic diversity of Blue 
Mountains Water Skink (Eulamprus leuraensis) 

##Abstract
Species can respond to global climate change in several different ways, either by persisting in 
their current environment, migrating, or going extinct. However, with species such as the 
Blue Mountains Water Skink (Eulamprus leuraensis) that are habitat specialists and restricted 
to montane habitats, migrating, or persisting in new environments may prove difficult. Using 
the dataset from Dubey and Shine (2010) consisting of 241 individuals and 13 populations, I
investigate the effects of mountain elevation on the genetic diversity of Eulamprus leuraensis
species. I find a significant relationship between allelic richness (AR) and mountain elevation 
(P=0.0022), with allelic richness increasing with higher elevation. I also find a significant 
relationship between expected heterozygosity (Hs) and elevation (P=0.0038), with expected 
heterozygosity increasing with higher elevation. These findings suggest restricted gene flow 
caused by genetic drift among populations, explaining higher genetic diversity at higher 
elevation sites than lower elevation sites. The significance of these findings can predict 
current and potential future trends to a population’s gene flow and spatial genetic structure 
under changing landscape and environmental variables. Knowledge of such trends can guide 
conservation managers in facilitating appropriate strategies in preventing climate-induced 
extinction of these endangered populations. 


##Introduction
Global climate change is ranked as the most significant threat to biodiversity (Radchuk et al. 
2019). Such impacts may well initiate fragmentation, resulting in reduced population size and 
limited dispersal, and thereby leading to a loss of genetic diversity and gene flow (Radchuk et 
al. 2019). However, accurately predicting how species and populations respond to climate 
change remains a difficult task as they can respond in several different ways, either by 
persisting in their current environment, migrating, or going extinct (Coretes 2016). But 
considering species that are already restricted to montane habitats, migrating may not be an 2
option (Razgour 2015). Consistent with that prediction, persistence in new environments may 
prove difficult for species such as reptiles and amphibians that have low thermal tolerances 
making them sensitive to climate change (Razgour 2015). 

The Blue Mountains water skink (Eulamprus leuraensis) is one example of such a taxa that 
exhibits low thermal tolerances and is restricted to montane habitats (Dubey et al. 2013; 
Gorisen et al. 2016). E. leuraensis is a medium-sized (up to 20cm in length) skink found in 
the family Scincidae (Dubey et al. 2013). The species are found only in 30 isolated swamps at 
560-1060m elevations across the Newnes Plateau and the Blue Mountains of New South 
Wales Australia. Because the species is restricted to montane areas of high elevations, it may 
put them at greater risk from global climate change (Dubey et al. 2013). Lower rainfall and 
higher temperatures are predicted at higher elevations, which can have negative consequences 
on the lizard’s habitat as reduced rainfall and higher temperatures can dry out the hanging 
swamps (Dubey and Shine, 2010). Because of the species restricted distribution and ongoing 
threats, they are classified as endangered under the Environmental Protection and 
Biodiversity Conservation Act 1999 and the Threatened Species Conservation Act 1995
(Dubey and Shine, 2010). 

Using the dataset from Dubey and Shine (2010) consisting of 241 individuals and 13 
populations, I aim to investigate the effects of mountain elevation on the genetic diversity of 
E. leuraensis species. Based on the information from the literature, I hypothesise that there 
will be a difference in the genetic diversity of E. leuraensis populations between mountain 
elevations. I predict that allelic richness will be lower in higher elevations than lower 
elevations. I also predict that expected heterozygosity will be lower in higher elevations than 
lower elevations.


##Methods
Data
In this study, I used the data obtained from Dubey and Shine (2010) study comprising of 
Eulamprus leuraensis individuals provided to me by Associate Professor Adam Stow. The 
data set consists of 241 individuals and 13 populations across the Blue Mountains and 
Newnes Plateau of New South Wales, Australia. 

Statistical Analyses
All statistical analyses were performed using the software package R version (3.1.0) (R 
Development Core Team 2015). Linear regression was used to assess the relationship 
between the independent variables (elevation: levels 500, 600, 700, 800, 900, 1000, 1100 
metres) and the dependent variables (genetic diversity: levels allelic richness and expected 
heterozygosity). The statistical null hypothesis for linear regression is that the linear 
relationship between Y (allelic richness and expected heterozygosity) and X (elevation) is 
equal to 0. The alternative hypothesis is that the linear relationship between Y (allelic 
richness and expected heterozygosity) and X (elevation) is not equal to 0. 
Assumptions of linear regression are independence of data within and among samples, X 
variable is measured without error, normality of the residuals about the fitted line, m X + b, 
homogeneity of variances for residuals about the fitted line m X + b, and a linear relationship. 
To ensure that this study includes an adequate number of replicates given logical constraints a 
post-hoc priori power analyses were undertaken on both measurement variables (allelic 
richness and expected heterozygosity). An arcsin square root-transformation was conducted 
on allelic richness (dependent) and elevation (independent) variables, to ensure the logtransformed data follow a log-normal distribution. However, expected heterozygosity 
(dependent) and elevation (independent) did not need to be log-transformed as they followed
a normal distribution.


##Results
There is a significant relationship between allelic richness (AR) and mountain elevation
(Linear regression: Df=1,11, R2=0.59, P=0.0022), with allelic richness increasing with higher 
elevation (Figure 1). Therefore, I reject the null hypothesis that the linear relationship 
between allelic richness and elevation is equal to 0. A 95% confidence interval for the slope 
is (+/- 0.111, 0.074). A post hoc priori power analysis on allelic richness (Power=0.90), 
showed that an adequate number of replicates were used in this experiment.


```{r Plot1, echo=FALSE}
```
View(Lizard_data)
plot(Lizard_data$Elevation,Lizard_data$AR, xlab='Elevation',ylab='AR',col='green')
plot(Lizard_data$Elevation,Lizard_data$AR, xlab='Elevation (m)',ylab='AR',col='green')
abline(Lizard_data$Elevation,Lizard_data$AR)
mymodel<-lm(Elevation~AR,data=Lizard_data)
mymodel<-lm(Elevation~AR,data=Lizard_data)
abline(mymodel)
anova(mymodel)
summary(mymodel)

##Figure 1 Caption
Figure 1: Relationship between elevation (m) and allelic richness (AR) across 13 populations
of E. leuraensis individuals in the Newnes Plateau and Blue Mountains of New South Wales, 
Australia. 


There is also a significant relationship between expected heterozygosity (Hs) and elevation
(Linear regression: Df=1,11, R2=0.54, P=0.0038) (Figure 2), with expected heterozygosity 
increasing with higher elevation. Therefore, I reject the null hypothesis that the linear 
relationship between expected heterozygosity (Hs) and elevation is equal to 0. A 95% 
confidence interval for the slope is (+/- 1174.62, 322.69). A post hoc priori power analysis on 
expected heterozygosity (Power=0.90), showed that an adequate number of replicates were 
used in this experiment.


```{r Plot2, echo=FALSE}
```
plot(Lizard_data$Elevation,Lizard_data$Hs, xlab='Elevation (m)',ylab='Hs',col='green')
abline(Lizard_data$Elevation,Lizard_data$Hs)
mymodel<-lm(Elevation~Hs,data=Lizard_data)
abline(mymodel)
anova(mymodel)
summary(mymodel)
plot(mymodel)


##Figure 2 Caption
Figure 2: Relationship between elevation (m) and expected heterozygosity (Hs) across 13 
populations of E. leuraensis individuals in the Newnes Plateau and Blue Mountains of New 
South Wales, Australia. 


Both plots illustrate an approximately linear relationship between allelic richness and 
elevation and expected heterozygosity and elevation, even though one outlier was detected in 
each dataset (Figure 3 and 4). This suggests that the assumption the relationship is linear is 
reasonable. It also illustrates that the variation around the estimated regression line is constant 
suggesting that the assumption of equal variances is reasonable. The assumption of normality 
also seems reasonable as values follow a reasonably straight line.

##Figure 3 Caption
Figure 3: Normal qqplot illustrating the relationship between elevation and allelic richness 
(AR). 

##Figure 4 Caption 
Figure 4: Normal qqplot illustrating the relationship between elevation and expected 
heterozygosity (Hs) 

##Discussion
Previously, I hypothesized that there will be a difference in genetic diversity of E. leuraensis 
populations between elevations. I predicted that both allelic richness (AR) and expected 
heterozygosity (Hs) will be lower in higher elevations than lower elevations. However, my 
results did not support my predictions and revealed that both allelic richness (AR) and 
expected heterozygosity (Hs) was higher in higher elevations than lower elevations across the 
13 populations. My results are consistent with the findings of Polato et al. (2017) in their 
study of montane mayflies revealing higher genetic diversity at higher elevations than lower 
elevations. The findings from Polato et al. (2017) study, as well as the findings from my 
research, suggests reduced gene flow separated by either genetic drift or topographical 
barriers amongst population sites. Indeed, E. leuraensis populations are restricted to 30 
isolated swamps separated by about 20km in two distinct patches (in the Newnes Plateau and 
the Blue Mountains) resulting in population fragmentation and thereby leading to reduced 
gene flow and genetic differentiation (Dubey and Shine, 2010). 

The reduced gene flow of remnant populations of E. leuraensis has considerable
consequences for the conservation of this endangered species (Coretest 2016; Cushman and 
Landguth, 2010). Because my study showed higher genetic diversity at high-elevation sites,
E. leuraensis populations are expected to be at higher risk of climate-induced extinction due 
to limited amounts of genetic variation and lower effective population size (Radchuk et al. 
2019). Climate variation models predict that E. leuranesis habitats will become drier (by up 
to 40%) and warmer (by up to 5 degrees C) within the next century (Razgour 2015). Such 
changes can have cascading consequences on the lizards and their montane habitats, via an 
increase in bushfire intensity and frequency, and their swamps drying out (Dubey and Shine, 
2010; Gorisen et al. 2016). Furthermore, E. leuranesis are also specialist species, making 
them more sensitive to environmental changes, thus their conservation remains an urgent 
priority (Dubey and Shine, 2010). 

As E. leuraensis are an endangered species, previous research has suggested habitat 
colonization as a potential conservation strategy (Gorisen et al. 2016). However, this method 
may not be possible as E. leuraensis populations inhabit high elevations, hence future work 
should consider habitat corridors as a potential alternative (Ramiadantsoa et al. 2015). The 
success of habitat corridors has been demonstrated in Ramiadantsoa et al. (2015) study 
revealing increased viability and landscape connectivity of fragmented populations in 8
Madagascar. Future work should also consider the effects of elevation on the genetic 
diversity of different species and taxa, including their physiological responses to different 
elevations (Segelbacher et al. 2010). One of the limitations of this study was that it only 
focused on the effects of mountain elevation on the genetic diversity of one species. 
Addressing the effects of elevational gradients in different species and taxa will be an 
important next step in further research into species vulnerability to elevational range shifts 
and climate change (Polato et al. 2017). 


##Acknowledgements 
I would like to thank Associate Professor Adam Stow and Dr Rachael Dudaniec for providing 
me with the data and their assistance with my Year 2 project. I would also like to thank Dr 
Drew Allen for their assistance with making the statistical models.


##References
Coretes, A.J. (2016) Local Scale Genetic Diversity and its Role in Coping with Changing 
Climate. Genetic Diversity. 1, 1-12. 
Cushman, S.A. and Landguth, E.L. (2010) Spurious correlations and inference in landscape 
genetics. Molecular Ecology. 19(1), 3592-3602. 
Dubey, S., Sinsch, U., Dehling, M.J., Chevalley, M. and Shine, R. (2013) Population 
demography of an endangered lizard, the Blue Mountains Water Skink. BMC Ecology. 13, 1-
12. 
Dubey, S. and Shine, R. (2010) Restricted dispersal and genetic diversity in populations of an 
endangered montane lizard (Eulamprus leuraensis, Scincidae). Molecular Ecology. 19, 886-
897. 
Gorissen, S., Greenlees, M. and Shine, R. (2016) A skink out of water: impacts of 
anthropogenic disturbance on an Endangered reptile in Australian highland swamps. Pacific 
Conservation Biology. 51, 1-16. 9
Olah, G., Smith, A.L., Asner, G.P., Brightsmith, D.J., Heinsohn, R.G. and Peakall, R. (2016) 
Exploring dispersal barriers using landscape genetic resistance modelling in scarlet macaws
of the Peruvian Amazon. Landscape Ecology. 32(1), 445-456. 
Polato et al. (2017) Genetic diversity and gene flow decline with elevation in montane 
mayflies. Heredity. 119, 107-116. 
Radchuk et al. (2019) Adaptive responses of animals to climate change are most likely 
insufficient. Nature Communications. 10, 1-14. 
Ramiadantsoa, R., Ovaskainen, O., Rybicki, J. and Hanski, I. (2015) Large-Scale Habitat 
Corridors for Biodiversity Conservation: A Forest Corridor in Madagascar. Plos One. 10(7), 
1-18. 
Razgour, O. (2015) Beyond species distribution modeling: A landscape genetics approach to 
investigating range shifts under future climate change. Ecological Informatics. 30, 250-256. 
R Core Team (2019). R: A language and environment for statistical computing. R Foundation 
for Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.
Segelbacher, G., Cusman, S.A., Epperson, B.K., Fortin, M.J., Francois, O., Hardy, O.J., 
Holderegger, R., Taberleet, P., Waits, L.P. and Manel, S. (2010) Applications of landscape 
genetics in conservation biology: concepts and challenges. Conservation Genetics. 11, 375-
385.


